---
title: "Slope and aspect creation with DEM"
---

I have a DEM file which is a tiff foramt type acquired for North Western Territory in Canada for the year 2023.

The name of the data is `NWT- 2023-DEM.tif` and it contains a raster image and source note for NWT- 2023-DEM.tif


To make it easy to share this dataset I will read it, convert it to an `.RData` file, and then document it so that it can be bundled with package `packr`.

Load the packages needed to preprocess the data:
```{r}
# install.packages("raster")
# library(raster)

```

Read the raster data:
```{r}
elevation_raster <- raster("/Users/chinyereottah/Desktop/surfaceR/data-raw/NWT-2023-DEM.tif")
```

Get a summary of the raster:
```{r}
summary(elevation_raster)
```


The source for these data are as follow:

- 2023 DEM of NWT, Canada :Chttps://search.earthdata.nasa.gov/search/?fi=ASTER


In addition, I want to fill the DEM. Filling a DEM when working with rasr data ensures that the  ensure that the elevation data is continuous and there are no missing values. 
```{r}
elevation_raster <- raster("/Users/chinyereottah/Desktop/surfaceR/data-raw/NWT-2023-DEM.tif")
# Check if there are any NAs (missing values) in the DEM
has_nas <- any(is.na(values(elevation_raster)))

if (has_nas) {
  # Fill the gaps using a simple interpolation method (linear interpolation)
  filled_dem <- fillRaster(elevation_raster, method = "bilinear")
} else {
  filled_dem <- elevation_raster
}
```

Save to data folder:
```{r}
writeRaster(elevation_raster, filename = "NWT-2023-DEMfill.tif", format = "GTiff")

```
