---
title: "surfaceR-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{surfaceR-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: "Chinyere Ottah"
date: "07 November 2023"
---

# Introduction
SurfaceR is a package for deriving slope and aspect using Digital Elevation Models (DEMs) in R. Slope and aspect influence fire behavior, making these variables crucial when assessing fire disturbances.

Steep slopes increase the spread of fire as fires tend to progress uphill because the heat and flames have already heated up the vegetation as the fire moves uphill, making it easier for them to burn and spread. Aspect can be used to determine the direction of the slope. South-facing slopes tend to have more sunlight, especially within the northern hemisphere, making the fuels drier and more susceptible to burning. Therefore, the surfaceR package is very useful for assessing fire disturbances, including emissions and other standing and non-standing disturbances.

This vignette provides an illustration of how surfaceR can be used for surface analysis following applications in the North Western Territories of Canada following a fire. From a provided digital elevation model for NWT (aster), surfaceR is used to calculate slope and aspect and extract values for statistical modeling and predictions following a fire.

# The surfaceR Package
The surfaceR package provides two terrain surface outputs:

- **Slope:** Indicates the steepness and inclination of a surface area.
- **Aspect:** Describes the direction of the slope and is measured in degrees (North, East, West, and South).

## Initializing the surfaceR Geoprocessing Environment
Before installing the surfaceR package, you need to make sure that you have the `raster` package. You can install it with the following commands:

```{r}
#install.packages("raster")
#library(raster)
```

#Installing the Package
Before using the calculate_slope_aspect function, you'll need to install and load the "surfaceR" package. You can do this using the following commands:

```{r}
# Install and load the 'surfaceR' package
devtools::install_github("ChinyereRuth/surfaceR")
library(surfaceR)
```

# Using calculate_slope_aspect
The calculate_slope_aspect function uses a DEM as its input and returns both the slope and aspect rasters. For instance, aspect and slope can be be derived from a DEM using the command line below.
First, read your DEM data:

```{r}
# Load the 'raster' package
library(raster)
elevation_raster <- raster("/Users/chinyereottah/Desktop/surfaceR/inst/extdata/NWT-2023-DEMfill.tif")
```
Where:

elevation_raster refers to the name of the object holding the raster data.
<- is the assignment operator in R.
raster() is a function from the raster package that is used to read the DEM file, which is a raster.
"/path/to/your/DEM.tif" refers to the path where the DEM data you want to read is.

#Calculate Slope and Aspect
The next step is to calculate slope and aspect using the calculate_slope_aspect function.

```{r}
# Load the 'surfaceR' package
library(surfaceR)
result <- calculate_slope_aspect(elevation_raster)

```
In the function above:

calculate_slope_aspect is the function that is called from the surfaceR package. This function can calculate aspect and slope using a DEM.
elevation_raster is the elevation data that is being passed as an argument to compute aspect and slope.

#Access the Slope and Aspect Rasters
The slope and aspect can be accessed using the following command:
```{r}
slope_raster <- result$slope
aspect_raster <- result$aspect

```
Where slope_raster is assigned values stored in result$slope, and aspect_raster is assigned values stored in result$aspect

# Visualizing the Results
The plot function can be used to visualize the results using the code below:

```{r}
# Plot the slope raster
plot(slope_raster, main = "Slope")

# Plot the aspect raster
plot(aspect_raster, main = "Aspect")

```

#Interpreting the Results
After plotting the results, we see the output for the slope and aspect raster. The slope_raster contains information about the slope for the area in degrees, while the aspect_raster contains values about the direction of the slope in degrees.

#Conclusion
With the calculate_slope_aspect function in surfaceR, slope and aspect can be easily computed just by using the DEM model as input. Information about slope and aspect is very important for fire modeling and carbon emission modeling improvements and accounting.

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(surfaceR)

```

